---
title: "Wrangling presidential election data"
author: "Jae Yeon Kim"
date: format(Sys.Date())
output: html_document
---

# Load packages 

```{r}

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
    tidyverse, 
    here
)

```

# Load data 

Huge thanks to [`tonmcg`](https://github.com/tonmcg) for putting this data together: https://github.com/tonmcg/US_County_Level_Election_Results_08-20

```{r}
election_08_16 <- read_csv("https://github.com/tonmcg/US_County_Level_Election_Results_08-20/raw/master/US_County_Level_Presidential_Results_08-16.csv")

election_20 <- read_csv("https://github.com/tonmcg/US_County_Level_Election_Results_08-20/raw/master/2020_US_County_Level_Presidential_Results.csv")
```

# Wrangle data 

## Find common columns

No common columns because of the column name inconsistency. 

```{r}
intersect(names(election_08_16), names(election_20))
```

Fix this issue by renaming the column names of `election_20`.

```{r}
# Check the column names 
names(election_08_16)
names(election_20)

# Rename common columns
election_20 <- election_20 %>%
    rename("fips_code" = "county_fips",
           "county" = "county_name")

# Check the common columns
intersect(names(election_08_16), names(election_20))
```

## Join the two datasets 

Caveat 1: Note that we have three more counties in the `election_20` dataframe than the `election_08_16` dataframe.

```{r}
length(unique(election_08_16$county))
length(unique(election_20$county))
```

Caveat 2: Also, 64 county names in the two dataframes didn't match.

```{r}
# The number of differently named counties 
setdiff(election_08_16$county, election_20$county) %>% length()

# Sort these counties 
setdiff(election_08_16$county, election_20$county) %>% sort() %>% head()
```

- Change the column names of the `election_20` in the same style of `election_08_16` dataframe. 

```{r}

election_20 <- election_20 %>%
    rename("dem_2020" = "votes_dem",
           "gop_2020" = "votes_gop",
           "total_2020" = "total_votes") %>%
    select(!matches("per_|diff"))

```

Since we're trying to create a panel data, I join the two datasets (x1, x2) based on x1. 

```{r}
joined <- election_08_16 %>% left_join(election_20) %>%
    # Also move the geo-related columns to the front 
    relocate(fips_code, county, state_name)
```

## Select

Let's focus on the 2016 and 2020 presidential election results. 

```{r}
selected <- joined %>%
    select(c("fips_code", "county", "state_name") | matches("2016|2020"))
```

## Mutate 

Calculate party vote share in each election and add that information to the original dataframe.

```{r}

augmented <- bind_cols(selected, 
          selected %>%
    transmute(dem_2016_share = dem_2016/total_2016, 
           gop_2016_share = gop_2016/total_2016, 
           dem_2020_share = dem_2020/total_2020, 
           gop_2020_share = gop_2020/total_2020) %>%
    transmute(dem_diff = (dem_2020_share - dem_2016_share) %>% round(4), 
              gop_diff = (gop_2020_share - gop_2016_share) %>% round(4))
)
  
head(augmented)
```

# Export 

```{r}
write_csv(augmented, here("outputs", "wrangled.csv"))
```

